A prototype software framework for transparent, reusable and updatable computational health economic models

### 2023-10-22 Reports/Manuscript/Manuscript.tex
problems with table and figure placement

  \usepackage{unicode-math}

@@

@@

@@

@@

\urlstyle{same} % disable monospaced font for URLs

@@

  pdftitle={A framework for implementing an open source and modular economic model of youth mental health that is accountable, reusable and updatable},
  pdfauthor={Matthew P Hamilton1,2,; Caroline X Gao1,3,2; Glen Wiesner4; Kate M Filia1,3; Jana M Menssink1,3; Petra Plencnerova5; David Baker1,3; Patrick D McGorry1,3; Alexandra Parker6; Jonathan Karnon7; Sue M Cotton1,3; Cathrine Mihalopoulos2},
  pdfkeywords={open source models, health economics, mental disorders},

@@

\title{A framework for implementing an open source and modular economic model of youth mental health that is accountable, reusable and updatable}
\author{Matthew P Hamilton\textsuperscript{1,2,*} \and Caroline X Gao\textsuperscript{1,3,2} \and Glen Wiesner\textsuperscript{4} \and Kate M Filia\textsuperscript{1,3} \and Jana M Menssink\textsuperscript{1,3} \and Petra Plencnerova\textsuperscript{5} \and David Baker\textsuperscript{1,3} \and Patrick D McGorry\textsuperscript{1,3} \and Alexandra Parker\textsuperscript{6} \and Jonathan Karnon\textsuperscript{7} \and Sue M Cotton\textsuperscript{1,3} \and Cathrine Mihalopoulos\textsuperscript{2}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
\textbf{Summary: } Compared to other disciplines, health economics rarely uses modular approaches to model development that facilitate re-use of model components. Although interest in open source health economic models (OSHEMs) is growing, uptake remains limited. We propose a framework for a modular OSHEM in youth mental health. The framework includes of set of X1 good practice standards for implementing a model that is accountable (X2 standards), reusable (X3 standards) and upatable (X4 standards). We provide a rationale for each standard. The framework also includes software toolkit of six R packages for authoring model modules (data structures and algorithms), supplying those modules with data and using the modules to implement reproducible modelling analyses. We describe an early application of the framework to developing utility mapping models that meet X5 standards. We discuss the potential benefits and challenges of extending this initial work to develop a more extensive model for undertaking economic analyses in youth mental health. \newline \newline \textbf{Code: } Visit \url{https://www.ready4-dev.com} for more information about how to find, install and apply the framework and model. \newline \newline
\end{abstract}

\textsuperscript{1} Orygen, Parkville, Australia\\
\textsuperscript{2} School of Public Health and Preventive Medicine, Monash University, Clayton, Australia\\
\textsuperscript{3} Centre for Youth Mental Health; The University of Melbourne, Parkville, Australia\\
\textsuperscript{4} Heart Foundation, Melbourne, Australia\\
\textsuperscript{5} headspace National Youth Mental Health Foundation, Melbourne, Australia\\
\textsuperscript{6} Victoria University, Footscray, Australia\\
\textsuperscript{7} Flinders University, Adelaide, Australia

\textsuperscript{*} Correspondence: \href{mailto:matthew.hamilton@orygen.org.au}{Matthew P Hamilton \textless{}\href{mailto:matthew.hamilton@orygen.org.au}{\nolinkurl{matthew.hamilton@orygen.org.au}}\textgreater{}}

@@

Computational models, particularly those addressing economic topics, have become essential tools for health policy development {[}1,2{]}. Although influential and widely used, health economic models typically have a number of limitations that at best restrict their usefulness and which in some cases may facilitate harmful misuse.

There is a strong case for making health economists more accountable for the appropriate use and social acceptability of their models. Users of models should be able to assess their adequacy for a particular purpose {[}3{]}. This goal is difficult to achieve in the current context of the poor reproducibility {[}4--6{]} and frequently insufficient validation {[}7{]} of health economic models. Value judgments strongly shape health economic analyses, yet are rarely made explicit, omissions that may lead to socially unacceptable policy recommendations {[}8{]}. A modelling team's value judgments about what questions to address, the most important features of a system to represent and the weighting of different types of evidence may be poorly aligned with those of the people impacted by decisions informed by model analyses {[}9{]}. As health economic models adopt more sophisticated techniques, the need for accountability grows. More complex models may be more prone to propagation errors {[}10{]} and models designed to address multiple questions should be expected to meet more onerous verification and validation obligations {[}11,12{]}. The nature and extent of individual model authorship contributions may be less clear in models implemented over longer time-frames with a large and changing group of collaborators {[}\textbf{thompson2022?}{]}.

Collaborative approaches that encourage re-use of models by other modellers can help make modelling pojects more tractable {[}13{]}. However, as many health economic models are owned by pharmaceutical companies and consultancies, commercial considerations can limit the reuse of models and their constituent code and data {[}12{]}. Not all data included in health economic models can be ethically made publicly available due to privacy and confidentiality requirements. Transferring a health economic model developed for one jurisdiction for application in another will involve retaining some features and updating others {[}14{]}. Use of concepts with standardised meanings across jurisdictions {[}15{]}, supplying context specific data as replaceable data-packs and distributing source code under licenses that allow derivative works are all steps that can enhance model transferability.

Health economic models should be updated and refined as new evidence emerges and decision contexts change {[}16{]}, but this occurs infrequently {[}17{]}. Decision makers are frequently left to rely on economic analyses that become less valid and relevant each year. A model can grow stale when changes in the real world, whether sudden (e.g.~new pandemics) or gradual (e.g.~demographic), is not represented by corresponding updates to relevant model features. Even when the system being modelled remains relatively stable, new research findings may potentially invalidate aspects of a model's original conceptualisation and implementation. Funding for health economic modelling projects rarely extend to provision of medium term support for model updates and improvements. The career trajectories of health economists can also mitigate against adequate maintenance of a model, it being relatively common for model authors to have moved on from the team that owns the model and / or from working on the health condition for which the model was developed.

A potential strategy for improving model accountability, reusability and updatability is to make models both modular and open-source.

A modular approach to model building involves models being constructed by combining multiple replaceable sub-models (modules) {[}18{]}. When used as model sub-components, modules communicate with other modules but they can be also run independently {[}19{]}. Advantages of modular models include feasibility (large projects are broken into smaller tasks, with each component independently developed and tested) and flexibility (making it easier to selectively replace or update specific parts of a model and to scale up or down the level of granularity) {[}18{]}. Modular approaches are currently being used to facilitate the development of complex computational models in disciplines such as biology {[}18{]}, neuroscience {[}20{]} and ecology {[}19{]}. Although the related and enabling concept of reusable reference models has been recommended in health economics {[}21{]}, peer reviewed studies describing modular health economic models remain relatively rare, though examples exist in tuberculosis {[}22{]} and cardiology {[}23{]}.

The benefits of modular models can be enhanced if multiple modelling teams have the capacity to contribute and use model modules. To facilitate this, modular models may be implemented as open source projects that grant liberal permissions to access and re-use model source code and data {[}18--20{]}. Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs) {[}24{]}, actual implementations are rare {[}12,25,26{]}. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations {[}24,27{]}. Adherence to good practice guidance is an essential requirement for healthcare modelling {[}2{]}, but guidelines for implementing OSHEMs remain scarce, piecemeal and need improving {[}28{]}.

We plan to develop a modular OSHEM of youth mental health. Open source approaches have been recommended to help develop the mental health modelling field {[}29{]} but only one mental health related model (in Alcohol Use Disorder {[}30{]}) is currently indexed {[}RECHECK{]} in the Open Source Models Clearinghouse {[}25,31{]}. Other than our own work, we are aware of just one open source mental health reference model - in Major Depressive Disorder - that is currently in development {[}32{]}.

In this paper, we introduce a framework for developing a modular youth mental health OSHEM. Specifically we describe the framework's:

\begin{itemize}
\item
  good practice standards for implementing a modular OSHEM;
\item
  software toolkit for developing a modular OSHEM consistent with those good practice standards; and
\item
  application to implementing model modules to support the development and use of utility mapping models.
\end{itemize}

\hypertarget{guidelines-for-oshems}{%
\section{Guidelines for OSHEMs}\label{guidelines-for-oshems}}

Guidelines on health economic model transparency were published ten years ago {[}11{]} and made recommendations on documenting models; however, notably missing were recommendations on the sharing of model code and data. More recent and more general modelling guidance {[}2{]} recommends the sharing of code and data through digital repository services and the use of version control systems. Based on these and other existing sources of guidance and our own experience, we propose 20 guidelines for implementing OSHEMs that are \textbf{TIMELY} (Transparent, Iterative, Modular, Epitomising, Licensed and Yielding - see Table @ref(tab:timelygls)).

@@

\caption{(\#tab:timelygls)TIMELY Guidelines for OSHEMs}
\centering
\begin{tabular}[t]{>{}ll>{\raggedright\arraybackslash}p{35em}}
\toprule
Characteristic & Guideline & Meaning\\
\midrule
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{Transparent - people can easily see how a model has been implemented and tested}}\\
\hspace{1em}\textbf{} & T1 & Uniquely identified copies of model code and data are permanently archived in open online repositories\\
\hspace{1em}\textbf{} & T2 & Model code and data are documented\\
\hspace{1em}\textbf{} & T3 & Model code uses a simple and consistent syntax\\
\hspace{1em}\textbf{} & T4 & Model analyses and reporting are implemented using literate programming\\
\hspace{1em}\textbf{} & T5 & Code coverage is reported\\
\hspace{1em}\textbf{} & T6 & All parts of a study analysis and reporting workflow can be reproduced and/or replicated\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{Iterative - a model is routinely updated to maintain and improve validity}}\\
\hspace{1em}\textbf{} & I1 & Model code and data are version controlled\\
\hspace{1em}\textbf{} & I2 & Model code and data use semantic versioning\\
\hspace{1em}\textbf{} & I3 & Continuous integration is used to verify model code updates\\
\hspace{1em}\textbf{} & I4 & Deprecation conventions are used to retire model code and data\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{Modular - models and their components can be combined to extend their scope}}\\
\hspace{1em}\textbf{} & M1 & Model code and data are stored and managed separately\\
\hspace{1em}\textbf{} & M2 & Model code defines encapsulating data structures\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{Epitomising - model code can be re-used in other decision contexts}}\\
\hspace{1em}\textbf{} & E1 & Model code is distributed as libraries of classes and functions\\
\hspace{1em}\textbf{} & E2 & Model code defines inheriting data-structures\\
\hspace{1em}\textbf{} & E3 & Test data is available to demonstrate generalised applications of model code\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{Licensed - a model and its components are persistently re-usable by other modellers}}\\
\hspace{1em}\textbf{} & L1 & Model code is made available for re-use under copyleft or permissive licenses\\
\hspace{1em}\textbf{} & L2 & Non-confidential model data is licensed for liberal re-use (subject to additional terms for de-identified human data)\\
\hspace{1em}\textbf{} & L3 & Model code and data are distributed with tools to support appropriate citation\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{Yielding - a model can be simply, flexibly and reliably used as a decision aid}}\\
\hspace{1em}\textbf{} & Y1 & Models are distributed with validated tools to support their safe and appropriate re-use\\
\hspace{1em}\textbf{} & Y2 & Simple user-interfaces allow non-technical users to configure and run models\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{transparent-models}{%
\subsection{Transparent models}\label{transparent-models}}

A range of tools and practices are available to help make models more transparent. The most efficient way to widely disseminate code and data may be to use existing open science infrastructure {[}2{]} to make permanently accessible and uniquely identfied digital artefacts. Model code and data also need to be clearly documented, potentially with different versions for technical and non-technical users {[}11{]}. Consistent use of meaningful naming conventions when authoring code is recommended {[}33,34{]}. Code can be made easier to follow by using the practices of abstraction {[}35{]}, where only simple, high level commands are routinely exposed to reviewers, and polymorphism {[}36{]}, where the same command (e.g.~``simulate'') can be reused to implement different algorithms of the same type. Programs to implement model analyses can be made comprehensible to even non-technical users through the use of literate programming techniques to render documents that integrate computer code with plain English descriptions.

An essential component of quality assuring health economic models is verification - ensuring that calculations are correct and consistent with model specifications {[}37{]}. One useful concept for informing model users about the extensiveness of verification checks is code coverage {[}38{]} - the proportion of model code that has been explicitly tested. Finally, transcription errors - mistakes introduced when transferring data between sources, models and reports - are very common in health economic models {[}39{]}. The risk of these errors might be lower if there was full transparency across all steps in a study workflow. Scientific computing tools now make it relatively straightforward to author programs that reproducibly execute all steps in data ingest, processing and reporting {[}33{]}.

\hypertarget{iterative-models}{%
\subsection{Iterative models}\label{iterative-models}}

To avoid OSHEMs going stale - losing validity and usefulness with time - they should be routinely updated. Each update of code and data should be uniquely identifiable and retrievable, a goal that can be facilitated by use of version control tools {[}2{]}.Labeling each change using semantic versioning {[}40{]} conventions can signal the potential importance of an update to users of model code and data. Continuous integration {[}41{]} tools can help verify that each code update passes multiple quality tests.
Finally, using deprecation conventions that take an informative and staged approach to retiring old code and data reduces the risk that model revisions have unintended consequences on third party users.\\

\hypertarget{modular-models}{%
\subsection{Modular models}\label{modular-models}}

An important consideration when combining model components (or modules) is to ensure that interactions between two modules do not compromise the validity of either. Using the coding practice of encapsulation {[}35{]} can help ensure that model modules can be safely combined {[}42{]}.

\hypertarget{epitomising-models}{%
\subsection{Epitomising models}\label{epitomising-models}}

A key challenge to generalising health economic models is that they are typically developed to inform a decision problem with a highly specific jurisdictional context. However, a number of choices about how these models are implemented can significantly increase the re-usability of model code in other contexts.

Writing code as collections of functions (short, self-contained and reusable algorithms that each perform a discrete task) is recommended as good practice for scientific computing {[}33{]}. When bundled for distribution as libraries, functions have the potential to be widely re-used, often in contexts very different than those they were originally developed for. A special type of function, called a method, can only be applied to a pre-defined class of data structure. Due to the coding concept of inheritance {[}35{]}, the more restricted nature of methods can be used to enhance the re-usability of model code in different decision contexts {[}42{]}. For example, when generalising a model developed for the Australian context to a UK context, one could create a class that initially inherits all of the methods defined for the Australian model and then write new or replacement methods as needed for the UK model.Whatever type of functions are written for a modelling project, it is good practice to make available test or toy data to demonstrate their use {[}33{]}.

\hypertarget{licensed-models}{%
\subsection{Licensed models}\label{licensed-models}}

To make model code and data widely re-usable by others, it is important to provide users with appropriate and explicit permissions. In the context of open source models, there are two broad categories of licensing options. Some guidance strongly recommends the use of permissive licensing {[}33{]} that provides users with great flexibility as to the purposes (including commercial) for which the content could be re-used. An alternative approach is to use copyleft licenses {[}43{]} that can require content users to distribute any derivative works they create under similar open source arrangements.The most suitable approach may differ for code and data.

For code, it may be appropriate to adopt the prevailing open source licensing practice within the programming language being used. For data, it may not be sufficient to simply choose between a permissive license like the Public Domain Dedication (CC0) {[}44{]} or a copyleft option such as the Attribution-Share Alike (CC-BY-SA) {[}45{]}. Responsible custodianship of some de-identified mental health data may involve using or adapting template terms of use {[}46{]} which have a number of ethical clauses (for example, prohibiting efforts to re-identify research participants). Licenses may or may not specify that model re-users must give appropriate acknowledgement to model authors. Citation tools can be distributed with each individual code or data item to inform re-users of the desired attribution.

\hypertarget{yielding-models}{%
\subsection{Yielding models}\label{yielding-models}}

OSHEMs can be time and skills intensive for modellers to develop - but they should be easy for others to use. Statistical models are a common output of health economic evaluations, but they are often not reported in a format that enables others to confidently and reliably re-use them {[}47{]}. Open source approaches can help address this by disseminating code artefacts that enable easy and appropriate use of a statistical model to make predictions with new data. However, great care must be exercised when doing so if models are derived from data on human subjects as some software artefacts by default contain a copy of the source dataset. Such dataset copies must therefore be replaced (for example, with synthetic data) and the amended artefact's predictive performance then retested before any public release. Another way to make OSHEMs easier to use is to develop simple user-interfaces for non-technical users.

\hypertarget{modelling-toolkit}{%
\section{Modelling Toolkit}\label{modelling-toolkit}}

We have developed a toolkit to help streamline the process of developing OSHEMs that meet the TIMELY standards. The toolkit is comprised of online repositories and software.

\hypertarget{repositories}{%
\subsection{\texorpdfstring{Repositories }{Repositories }}\label{repositories}}

We created a GitHub organisation where all framework software source code is stored, documented, version controlled and continuously integrated {[}48{]}. To store citable archived copies of release copies of our software, we created a Zenodo community {[}49{]}. Finally, to manage datasets for use in models developed with the framework, we created a dedicated collection within the Harvard Dataverse {[}50{]}.

\hypertarget{software}{%
\subsection{\texorpdfstring{Software }{Software }}\label{software}}

A coding framework for OSHEMs developed in the language R includes standardised approaches to directory structure and naming conventions {[}34{]}.

As a foundation for implementing the framework, we authored a development version R package that defines a novel syntax and a template class for model modules {[}51{]}. To enable the syntax and module template be applied to modelling projects, we then created five additional development version R packages that provide tools for authoring:

\begin{itemize}
\item
  documented model modules {[}52{]};
\item
  documented functions (including methods), written in a consistent house style {[}53{]};
\item
  citable, quality assured R packages {[}54{]};
\item
  model datasets {[}55{]};
\item
  model analyses and reports {[}56{]}.
\end{itemize}

The six R packages, their primary focus, the TIMELY standards they support and the third-party packages they depend on are summarised in Table @ref(tab:cpkgs). When used in conjunction with framework repositories, the six packages extend existing R packages provide strong support for implementing 16 of the TIMELY standards. However, the software only weakly supports implementing the standards relating to disseminating statistical models (Y1) and user-interface development (Y2) and does not yet provide any workflow tools to help implement the standards for code coverage (T5) and deprecation conventions (I4). Standards not supported or weakly supported by our software can be met with existing developer tools in R and we plan to progressively integrate these third-party tools with our own in future releases of our software. Another future priority is to submit production versions of each R package for review by and archiving on the Comprehensive R Archive Network (CRAN) {[}57{]}.

>>>>

  \usepackage{unicode-math} % this also loads fontspec

@@

\ifPDFTeX\else
  % xetex/luatex font selection

@@

\usepackage{pdflscape}

@@

\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{wrapfig}

@@

\urlstyle{same}

@@

  pdftitle={A prototype software framework for transparent, reusable and updatable computational health economic models},
  pdfauthor={Matthew P Hamilton1,2,; Caroline X Gao2,3,1; Glen Wiesner4; Kate M Filia2,3; Jana M Menssink2,3; Petra Plencnerova5; David Baker2,3; Patrick D McGorry2,3; Alexandra Parker6,3; Jonathan Karnon7; Sue M Cotton2,3; Cathrine Mihalopoulos1},
  pdfkeywords={computational models, ethics of modelling, health economics, mental disorders, open-source models, software frameworks},

@@

\title{A prototype software framework for transparent, reusable and updatable computational health economic models}
\author{Matthew P Hamilton\textsuperscript{1,2,*} \and Caroline X Gao\textsuperscript{2,3,1} \and Glen Wiesner\textsuperscript{4} \and Kate M Filia\textsuperscript{2,3} \and Jana M Menssink\textsuperscript{2,3} \and Petra Plencnerova\textsuperscript{5} \and David Baker\textsuperscript{2,3} \and Patrick D McGorry\textsuperscript{2,3} \and Alexandra Parker\textsuperscript{6,3} \and Jonathan Karnon\textsuperscript{7} \and Sue M Cotton\textsuperscript{2,3} \and Cathrine Mihalopoulos\textsuperscript{1}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
\textbf{Summary: } Most health economic analyses are undertaken with the aid of computers. However, the ethical dimensions of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. We propose that developers and funders of CHEMs share ethical responsibilities to (i) establish socially acceptable user requirements and design specifications; (ii) ensure fitness for purpose; and (iii) support socially beneficial use. We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely fulfilled these responsibilities. We propose six criteria for assessing CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; (R1) programming practices promote generalisability and transferability; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and (U2) new releases are systematically retested and appropriately deprecated. To facilitate CHEMs that meet TRU criteria, we have developed a prototype software framework in the open-source programming language R. The framework comprises six code libraries for authoring CHEMs, supplying CHEMs with data and undertaking analyses with CHEMs. The prototype software framework integrates with services for software development and research data archiving. We determine that an initial set of youth mental health CHEMs we developed with the prototype software framework wholly meet criteria T1-2, R1-2 and U1 and partially meet criterion U2. Our assessment criteria and prototype software framework can help inform and improve ethical implementation of CHEMs. Resource barriers to ethical CHEM practice should be addressed by research funders. \newline \newline \textbf{Code: } Visit \url{https://www.ready4-dev.com} for more information about how to find, install and apply the prototype software framework and CHEMs developed with it. \newline \newline
\end{abstract}

\textsuperscript{1} School of Public Health and Preventive Medicine, Monash University, Clayton, Australia\\
\textsuperscript{2} Orygen, Parkville, Australia\\
\textsuperscript{3} Centre for Youth Mental Health; The University of Melbourne, Parkville, Australia\\
\textsuperscript{4} Heart Foundation, Melbourne, Australia\\
\textsuperscript{5} headspace National Youth Mental Health Foundation, Melbourne, Australia\\
\textsuperscript{6} Institute for Health and Sport, Victoria University, Footscray, Australia\\
\textsuperscript{7} Flinders University, Adelaide, Australia

\textsuperscript{*} Correspondence: \href{mailto:matthew.hamilton1@monash.edu}{Matthew P Hamilton \textless{}\href{mailto:matthew.hamilton1@monash.edu}{\nolinkurl{matthew.hamilton1@monash.edu}}\textgreater{}}

@@

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services {[}1{]}. In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model). Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.

The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM). CHEMs can be developed using commercial modelling software or open-source programming languages such as Python {[}2{]} or R {[}3{]}. An advantage of commercial modelling software is that users often require limited or no programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable {[}4,5{]}. The decision about what type of software development platform to use must balance ethical considerations and resource constraints (see Table \ref{tab:proscons}).

@@

\caption{\label{tab:proscons}Implications of choosing commercial modelling software or an open-source programming language for the computational implementation of health economic models}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{7em}>{\raggedright\arraybackslash}p{20em}>{\raggedright\arraybackslash}p{20em}}
\toprule
Implications & Commercial modelling software & Open-source programming language\\
\midrule
 & (+) Will include tools to efficienty accomplish many common modelling tasks. & (-) Initial development is likely to be time and resource intensive.\\

 & (+) Requires limited or no programming skills. & (+) Over medium term, efficiency savings are possible if:\\

 & (-) Requires model developers and users to purchase software licenses & (i)  artefacts (code and data) from pre-existing models are leveraged; and / or\\

\multirow{-4}{7em}{\raggedright\arraybackslash \textbf{Resources}} &  & (ii)    maintaining and transferring (e.g. to new decision problems / contexts) model.\\
\cmidrule{1-3}
\addlinespace[0.3em]
\multicolumn{3}{c}{\textbf{Transparency}}\\
\hspace{1em} & (+) Popular tools (e.g. Excel) are readily understood by many. & (+) Easy to publicly share testable source code while selectively restricting access to confidential data.\\

\hspace{1em} & (-) Non seemless integration with scientific manuscript authoring pipelines raises risk of (undetected) transcription errors when reporting results. & (+) Easy to make clear who has developed and tested each part.\\

\hspace{1em} &  & (-) Novel code increases likelihood of (undetected) software errors.\\

\addlinespace[0.3em]
\multicolumn{3}{c}{\textbf{Reusability}}\\
\hspace{1em} & (+) Files will open and execute correctly for years after project is completed. & (-) Potentially fragile – if not maintained / not bundled with all required and correctly versioned dependencies\\

\hspace{1em} &  & (+) Facilitates modular implementations that make complex model representations more tractable.\\

\hspace{1em} &  & (+) Facilitates customised integration with existing data / decision systems.\\

\addlinespace[0.3em]
\multicolumn{3}{c}{\textbf{Updatability}}\\
\hspace{1em}\multirow{-7}{7em}{\raggedright\arraybackslash \textbf{Ethical}} & (-) Selective modification with appropriate attribution can be difficult. & (-) Facilitates collaborative development and maintenance.\\
\bottomrule
\end{tabular}
\end{table}

Open-source CHEMs remain rare {[}6,7{]}. The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists {[}5{]}. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations {[}5{]}.

Better software frameworks could potentially help address some of the barriers to open-source CHEM development. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users {[}8{]}. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers {[}9{]}. Software frameworks have widely been developed and implemented in data science, for example, PyTorch {[}10{]} for machine learning using Python.

A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency {[}11{]}. However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, the ethical responsibilities of public health computational modellers remain poorly understood {[}12{]} and the computational modelling landscape is changing fast {[}13{]}. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time {[}9{]}. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training {[}8{]}.

In this paper, we:

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\item
  propose a set of ethical responsibilities for CHEM project teams and criteria for assessing ethical CHEM implementations;
\item
  describe a prototype software framework we have developed to support the ethical implementation of CHEMs; and
\item
  assess an initial set of youth mental health CHEMs developed with our software framework against our proposed ethical CHEM implementation criteria.
\end{enumerate}

\hypertarget{ethical-issues}{%
\section{Ethical issues}\label{ethical-issues}}

We considered literature on modelling practice and reflected on our prior experience with developing and using CHEMs to identify: (i) some core ethical responsibilities of CHEM project teams; (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.

\hypertarget{ethical-responsibilities-of-chem-project-teams}{%
\subsection{Ethical responsibilities of CHEM project teams}\label{ethical-responsibilities-of-chem-project-teams}}

Boden and McKendrick {[}12{]} propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use). However, these criteria are not specific to the computational model, but are designed to apply to a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only ``is the model code open source or available on request'' (for transparency) and ``has the model been verified, i.e., does it do what the modeller wants it to do?'' (for beneficence / non-maleficence) are specific to model's representation as software.

To identify ethical issues that are specific to CHEMs, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases {[}14{]}. However, SDLC models typically have components that can principally map to the concepts of \emph{planning} (e.g., identification of system / user requirements and design specifications), \emph{implementation} (e.g., development and testing) and \emph{release} (e.g., deployment, system integration, maintenance and support). For simplicity, we propose one overarching ethical responsibility for CHEM project teams for each phase. These responsibilities are to:

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\item
  \textbf{establish socially acceptable user requirements and design specifications} (during CHEM planning);
\item
  \textbf{ensure fitness for purpose} (of CHEM implementation); and
\item
  \textbf{support socially beneficial use} (on CHEM release).
\end{enumerate}

Transparency underpins all three responsibilities (Table \ref{tab:timelygls}). Boden and McKendrick's other ethical criteria can be selectively mapped to each phase's responsibility: justice and independence to the planning responsibility; independence to the implementation responsibility; and beneficence / non-maleficence and justice to the release responsibility. All three ethical responsibilities have implications for both professional practice and project resourcing. As it is unreasonable to place expectations of modellers that they are not resourced to fulfil, these responsibilities should apply jointly to health economic model developers and funders.

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks {[}15,16{]}. The value judgments of health economic modellers are rarely adequately specified, omissions that may lead to socially unacceptable policy recommendations {[}16{]}. These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects {[}17{]}. For example, as illustrated in Table \ref{tab:proscons}, project resource constraints may require model developers to make trade-offs when agreeing CHEM \textbf{user requirements and design specifications}. These trade-offs in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \ref{tab:timelygls}).

Modellers have duties both to ensure a computational model is \textbf{fit for purpose} and to provide potential third-party users with the means of assessing its adequacy for their proposed uses {[}6,12,18,19{]}. However, it is common for health economic models to have serious methodological flaws {[}20,21{]}; insufficient validation {[}22--24{]}, poor reproducibility {[}25--27{]}; and undeclared errors {[}28{]}. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

A CHEM will have limited \textbf{social benefit} if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay. Reuse of CHEMs as components of other models can potentially make model development more efficient {[}29,30{]}. However, health economic models face challenges related to transferability across jurisdictions {[}30{]} that create barriers to reuse. Without ongoing maintenance, a CHEM risks becoming less reliable with time {[}30{]} and is at risk of being deployed for purposes for which it is poorly suited {[}13{]}. Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates {[}31{]}, thus limiting the temporal window within which a CHEM can be validly applied.

\hypertarget{chem-attributes-associated-with-ethical-modelling-practice}{%
\subsection{CHEM attributes associated with ethical modelling practice}\label{chem-attributes-associated-with-ethical-modelling-practice}}

The responsibilities of model project teams over a CHEM lifecycle are easier to state than to measure. However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs. As described in Table \ref{tab:timelygls}, we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations. We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{1.20in}|p{1.60in}|p{1.60in}|p{1.60in}}

\caption{Transparent,\ reusable\ and\ updatable\ (TRU)\ computational\ health\ economic\ models\ (CHEMs)\ are\ suggestive\ of\ ethical\ modelling\ practice}\label{tab:timelygls}\\

\ascline{1pt}{666666}{1-4}

\multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\centering}m{\dimexpr 1.2in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{\ }}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Social\ acceptability}}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Fitness\ for\ purpose}}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}!{\color[HTML]{666666}\vrule width 1pt}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Beneficial\ use}}}} \\

\ascline{1pt}{666666}{1-4}\endfirsthead \caption[]{Transparent,\ reusable\ and\ updatable\ (TRU)\ computational\ health\ economic\ models\ (CHEMs)\ are\ suggestive\ of\ ethical\ modelling\ practice}\label{tab:timelygls}\\

\ascline{1pt}{666666}{1-4}
